---
title: "Population prescribed drugs for anxiety or depression or psychosis"
author: "Fionnuala Cousins 52319277"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
knit: (function(inputFile, encoding) {
      out_dir <- "outputs";
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=file.path(dirname(inputFile), out_dir))})
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading packages}
# tidyverse includes dplyr and ggplot2 so I don't need to load them separately
library(tidyverse)
library(here)
library(viridis)
library(plotly)
```

```{r Step 0: Define Question}
```

# Question

What are the regional trends for the number of people prescribed drugs for 
anxiety, depression and psychosis in Scotland over the last ten years?
What might these mean for employers' allocation of support resources? The next 
sections follow the Health Data Science Workflow to address these questions.

# Data Acquisition

The data was downloaded from 
<https://scotland.shinyapps.io/ScotPHO_profiles_tool/> on 05/11/23 for the item
"population prescribed drugs for anxiety/depression/psychosis" for all available
years and all health boards. The downloaded file was called timetrend_data.csv


```{r Step 1: Data Acquisition}
#reading in the data and checking its columns:
adp_data <- read_csv(here("Inputs/timetrend_data.csv"))
glimpse(adp_data)
```

#Prepare/Clean Data

##Is it tidy? 

###Does each variable form a column? Yes, for example, Year is a variable so has a single column rather than a column for each year. 

###Does each observation form a row? Yes. While there are multiple columns, they are each just clarifying the observed value rather than additional observations themselves (for example, confidence intervals).

###Is each cell a single value? Yes, for example, there are no commas separating values. The period column data is in the format "YYYY/YYYY financial year" but this is still a single value because financial years always span two sequential calendar years. 

### Are there duplicate values?
```{r}
# checking for duplicates
unique_columns <- c("area_code", "year")
duplicates <- adp_data[duplicated(adp_data[, unique_columns]) | duplicated(adp_data[, unique_columns], fromLast = TRUE), ]
print(duplicates)
```

```{r Step 2: Prepare/Clear}
# 
# This chunk is for selecting and renaming columns and removing the NHS prefix.

clean_data <- adp_data %>%
  select('area_name','year','numerator', 'measure') %>% 
  rename(no_prescriptions = 'numerator', NHS = 'area_name', percentage_pop = 'measure') %>% 
  mutate(NHS = sub("^NHS ","", NHS)) %>% 
  mutate(board_population = no_prescriptions / percentage_pop * 100)
head(clean_data)

```
```{r}
# The below code plots the data as-is. It's here so anyone can plot it and see  
# for themselves that 14 is too many categories, but it's commented out so it 
# doesn't get confused with the intended plot.

# options(scipen = 999)
# clean_data %>% 
#  filter(NHS != "Scotland") %>% 
#   ggplot() +
  
#   geom_area(aes(x = year, y = no_prescriptions, fill = NHS))
```


#Analyse 

14 Health Boards are too many to plot in the same visualisation; the audience
would be overwhelmed. So, I will create a new column, classifying neighbouring
NHS boards into Central Belt, Borders, Highlands and Islands and North East. 
Then, I have to sum the old NHS Board percentages for each year into a single 
value for the Region for that year.

```{r Step 3: Analyse}
#I wanted to use functions from the course but when I looked up how to recategorise a categorical variable, all the answers were using functions we hadn't used. So, I asked chatgpt. First, it advised a statement for each NHS Board but I knew it could be done in groups and using | or %in% so I modified the prompt. 
group_data <- clean_data %>% 
  mutate(Region = case_when(
    NHS %in% c("Ayrshire & Arran" , "Borders" , "Dumfries & Galloway")
    ~ "Borders",
    NHS %in% c("Fife" , "Forth Valley" , "Greater Glasgow & Clyde" , 
               "Lanarkshire" , "Lothian")
    ~ "Central Belt",
    NHS %in% c("Grampian" , "Tayside")
    ~ "North East",
    NHS %in% c("Highland" , "Western Isles" , "Orkney" , "Shetland")
    ~ "Highlands & Islands"))


```


```{r}
#Now to re-calculate the regional percentages
summed_data <- group_data %>% 
  group_by(Region, year) %>% 
  mutate(region_pop = sum(board_population)) %>% 
  mutate(region_prescriptions = sum(no_prescriptions)) %>% 
  summarise(percentage = region_prescriptions / sum() * 100)
glimpse(summed_data)

```


```{r Step 4: Communication, fig.width = 10}
options(scipen = 999)
summed_data %>% 
  ggplot(aes(x = year, y = percentage, fill = Region)) +
  geom_area()+
  xlab("Year")+
  ylab("Percentage of people")+
  ggtitle("Trend in estimated population prescribed drugs for anxiety, depression or psychosis in Scotland by region")+
  scale_fill_viridis(discrete=TRUE)+
  scale_x_continuous(breaks = unique(summed_data$year))+
  theme_bw()

# If switching the output to html, this line shows values when hovered over. 
# It's commented out here because the output is PDF. 
# ggplotly(tooltip = c("x", "y"))

```

